---
title: "Peer-graded Assignment: Course Project 1"
output: 
  html_document:
    keep_md: yes
---

## Loading the libraries
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(lattice)
```

## Loading and preprocessing the data

Load the data from file `activity.csv`

```{r}
data <- read_csv('activity.csv')
```

We don't need any preprocessing now, but we take a look at the data we got.

```{r}
str(data)
```

As we see, some values are missing for steps.

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day

```{r}
stepsByDate <-
        group_by(data, date) %>%
        summarize(steps = sum(steps, na.rm=TRUE))
```

Make a histogram of the total number of steps taken each day
```{r}
hist(stepsByDate$steps, breaks=nrow(stepsByDate), main="Histogram of daily total number of steps")
```

Calculate and report the mean and median of the total number of steps taken per day
```{r}
summarize(stepsByDate,
        mean = mean(steps, na.rm = TRUE),
        median = median(steps, na.rm = TRUE))
```

## What is the average daily activity pattern?

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
meanStepsBy5minIntervals <-
        group_by(data, interval) %>% 
        summarize(steps=mean(steps, na.rm=TRUE))

plot(meanStepsBy5minIntervals, type="l")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxMeanStepsBy5minIntervals <- arrange(meanStepsBy5minIntervals, desc(steps))

maxMeanStepsBy5minIntervals[1,]
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset

```{r}
missingItemsCount <- sum(is.na(data$steps))
missingItemsCount
```

Fill in all of the missing values in the dataset. Replace each missing value by the mean number of steps at that interval in all days. We create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r}
fixedData <-
        mutate(data, steps = ifelse(is.na(steps), meanStepsBy5minIntervals$steps, steps))

fixedStepsByDate <-
        group_by(fixedData, date) %>%
        summarize(steps = sum(steps))
```

Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

```{r}
hist(fixedStepsByDate$steps, breaks=nrow(fixedStepsByDate),
     main="Histogram of daily total number of steps (fixed)")

summarize(fixedStepsByDate,
          mean = mean(steps),
          median = median(steps))
```

As we see, these values differ from those of the original data, repeated here:

```{r}
summarize(stepsByDate,
        mean = mean(steps, na.rm = TRUE),
        median = median(steps, na.rm = TRUE))
```

### Explaining the difference
Can we explain the difference?

Because we only add new values to `fixedStepsByDate`, it is clear that the sum of the total numbers of daily steps must be larger in `fixedStepsByDate` than in `stepsByDate`. The data frames `stepsByDate` and `fixedStepsByDate` have the same number of entries, one per each of the 61 different days.

```{r}
c(nrow(stepsByDate),nrow(fixedStepsByDate))
```

Therefore the mean of the total numbers of daily steps must be larger `fixedStepsByDate` than in `stepsByDate`

Why is the median and mean are exactly the same in `fixedStepsByDate`?

In the original data, if data was missing for one interval in a specific day, it was missing for the whole day (288 intervals). There were eight such days.

```{r}
group_by(data, date) %>% 
        summarize(misscount = sum(is.na(steps))) %>%
        filter(misscount != 0)
```

In `fixedStepsByDate`, the daily sums that were zero in `stepsByDate` are changed to the sum of the means of the steps in the different intervals, which is

```{r}
sum(meanStepsBy5minIntervals$steps)
```

This is also the mean of daily steps in the original data, if we drop all the dates that have missing values.

```{r}
filter(data, !is.na(steps)) %>%
        group_by(date) %>%
        summarize(totalSteps=sum(steps)) %>%
        summarize(mean=mean(totalSteps))
```

In `fixedStepsByDate`, we replaced the zero entries by this mean, and when we calculate the median, one of these values falls to the middle of the sorted vector and is therefore the median:

```{r}
sort(fixedStepsByDate$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
data <- mutate(data, daytype = factor(ifelse (weekdays(as.Date(date)) %in% c("Saturday", "Sunday"),
                                      "weekend", "weekday")))
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
xyplot(steps ~ interval | daytype, type='l', layout = c(1, 2),
        ylab="Number of steps", xlab="Interval",
        data = group_by(data, interval, daytype) %>%
                     summarize(steps=mean(steps, na.rm=TRUE)))

```